Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines

### 2013-03-15 paper/paper.tex
Starting rewrite.

\title{Optimizing Abstract Abstract Machines}

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all stores after stepping.

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) transition elimination
and (2) pragmatic improvement.
%
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
 \item lazy non-determinism;
 \item abstract compilation; and
 \item uniform literal approximation.
\end{enumerate}
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item store deltas;
 \item timestamped stores; and
 \item preallocated data structures.
\end{enumerate}

Some techniques preserve the precision of the underlying analysis, and others
do not.
%
For any technique that loses precision, we will discuss the design rationale
for the move.

@@

\begin{align*}
\force &: \Store \times \Value \to \Set(\Value) \\
\force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
\force(\msto,\mval) &= \{\mval\} \\
\ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto \force(\msto, \mval)] \\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
\mathit{bind\,}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \ext(\msto,\maddr, \mval), \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

@@

\paragraph{Option 1: Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step. With laziness, the
reference will mean the original binding(s) of the variable or the new
one, because the actual store lookup is delayed one step
(i.e. laziness is administrative). Without laziness, the reference
will fan out to all the bindings of the variable before the new
binding happens and thus might have an observable precision
difference.

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by duplicating the reduction relation to specialize
variable lookup. This works since in the semantics of ISWIM with
store-allocated results consumes the value component of states in one
step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligible benefit, we decided against this
approach.

@@

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \compile{\mexp}(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{align*}
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr},\mval,\msto} & \machstep
\mcomp^\mcntr(\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr})
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \machstep
\mcompi0^\mcntr(\menv,\msto,\mkont)
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \machstep
\mcompi1^\mcntr(\menv,\msto,\mkont)
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \machstep
\mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
\text{ where }\menv',\msto',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
\\
\ap{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto} \\
\text{ where }\mkont &\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\end{gather*}
\caption{Abstract abstract machine for compiled ISWIM}
\label{fig:caam}
\end{figure}

@@

Every step the analysis makes for the above techniques requires
joining large stores together. Not every step will modify all
addresses of the store, so joining entire stores is wasteful in terms
of memory and time. We can instead log store changes and replay the
change log on the full store after all steps have completed. This uses
far fewer join operations, leading to less overhead, and is \emph{precision-preserving}.

We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
 becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:
% TODO: Turn the recursion here into ...:
\begin{align*}
\replay &: \StoreDelta \times \Store \to \Store \\
\replay(\mtlst, \msto) &= \msto \\
\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
\end{align*}

@@

\begin{align*}
(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
\end{align*}
\begin{align*}
\mathit{bind\,}^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\msdiff}, \mcntr' \\
\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
               \maddr &= \mvar\mcntr'
\end{align*}
Compilation changes to additionally take a $\msdiff$ component, so the
above rule's right hand side would instead be $k^{\mcntr'}(\menv',
\msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.

We also lift $\machstep$ to accommodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated:

\begin{align*}
(\states, \msto) &\mathrel{\widehat{\machstep}} (\states\cup \states', \replay(\msdiff,\msto)) \\
\mbox{ where } (\states', \msdiff) &= \step^*(\varnothing, \states, \mtlst) \\
\step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
\step^*(S, \{c\}\cup \states, \msdiff &= \step^*(S\cup cs^*, \states, \msdiff^*) \\
\states^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
\msdiff^* &= \concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
\end{align*}

Here $\concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
of changes to one; the order in which it appends does not matter:
\begin{align*}
\concat(\varnothing) &= \mtlst \\
\concat(\{\msdiff\}\cup D) &= \append(\msdiff, \concat(D))
\end{align*}

@@

\subsection{Abstracting literal compound data}

The abstraction of compound data structures like lists has deep implications 
for precision and performance.
%
The classic list-heavy benchmark {\tt boyer}---a simplified implementation
of the Boyer-Moore theorem-prover---drove our own considerations for
abstractions for compound data structures.
%
Boyer's fluent use of literal list dooms the ``natural'' abstraction
of lists to fail from over precision.
%
In short, if we interpret a literal list {\tt '(a b c)} as {\tt (cons 'a (cons 'b
(cons 'c '())))}, even a monovariant allocation strategy for abstract
cons cells will precisely and exactly create a 3-cell list.

If code contains a length-300 literal list, then its analysis yields a length-300 abstract
list as well.
%
In many cases, the specific contents of the list add little precision
to the resulting analysis, and yet the analyzer will dutifully execute
recursive functions over the entirety of these structures at every
encounter.
%
An alternative in this case is to explicitly flatten literal lists
into single abstract cells.

We explore both options here.
%
We will first explain the natural abstraction of tuples, and then we
will explain a less precise allocation strategy from uniform $k$-CFA
that we use for large compound data literals that we modified from the
implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.

\subsubsection{Option 1: The natural abstraction}

The uniform way AAM approaches a simple abstraction strategy is to cut
recursion out of the data definition by tying the recursive knot
through the abstract store. For Scheme, the grammar for values looks
like the following:

\newcommand{\alt}{\;\mid\;}
\begin{align*}
\Value &::= \mbox{{\tt \#t}} \alt \mbox{{\tt \#f}} \alt \mbox{{\tt(cons $\Value$ $\Value$)}} \alt \mbox{{\tt '()}} \alt ...
\end{align*}

Upon evaluating a {\tt cons} application, we instead allocate two
addresses $a$ and $d$, join them to the respective values in the
store, and return the flattened {\tt (cons $a$ $d$)} value. Since
these addresses are all distinguished at different syntactic call sites
in the uniform $k$-CFA allocation strategy, and quoted lists are sugar
for a sequence of calls to {\tt cons}, this abstraction explodes the
value space. Analyzing a function that counts the number of atoms in a
literal s-expression would actually interpret that function at least
that number of times (more because of intermediate conses). Indeed,
even in our fastest implementation, this abstraction causes the
analysis of Boyer to be 430 times slower than the approach we will now
describe.

\subsubsection{Option 2: A precise yet compact abstraction}

The number of syntactic uses of {\tt cons} versus implicit uses via
literal lists is smaller in typical Scheme programs. We use the above
abstraction for these syntactic uses, but choose to interpret literal
lists as not always sugar for cascading {\tt cons}es. In particular,
if a list literal is ``too big'' (in our case length $>$ 1), we
interpret the list as a circular data structure; the right address
points back to the {\tt cons} itself, and the left address points to
all of the elements of the list. We take this farther and join the
list elements together in with a coarse type-based abstraction. In
effect, large lists or vectors of literal numbers become unbounded
lists/vectors of ``number.'' Heterogeneous data is combined to just
``data,'' rather than a union of ``number'', ``string'', etc. since
primitives are interpreted on every combination of values that flow to
them, and unions lead to more primitive interpretations.

Quotation is special because it cannot introduce function values,
which is important for a simple semantics. If we join two values of
different type together, we don't get ``anything,'' which has complex
meaning in higher-order languages (Shivers' escape semantics) and is
overly approximate. We instead get ``any quotable value,'' which has
much simpler semantics.

There are a few steps to consider:
\begin{itemize}
 \item{Define special value lattice elements for compound data domains that can be quoted
       (e.g. {\tt QPair} for $\left\{ ({\tt cons}\ a_i\ d_i)\right\}_i$, {\tt QVector} for immutable vectors, etc.)}
 \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
 \item{Interpret {\tt (quote (a ...))} as {\tt (qlist a ...)}, a new primitive function defined in figure \ref{fig:qlist},
       and similar definitions for immutable vectors.}
 \item{Extend the $\interpdelta$ axioms to include conservative
   meaning for these new values (e.g. {\tt (car QData)} = {\tt QData},
   {\tt (add1 QData)} = {\tt Number} and log ``possible type error'')
   and allow them to allocate addresses and change the store}
\end{itemize}

\begin{figure}
\begin{align*}
\interpdelta(\msto,\ {\tt qlist}) &= (\mbox{{\tt '()}}, \msto) \\
\interpdelta(\msto,\ {\tt qlist}, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
                            &\qquad \sqcup[d \mapsto \{ \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\}] \\
\bigsqcup(v) &= v \\
\bigsqcup(v, vs ...) &= \fmerge(v, \bigsqcup(vs ...)) \\
\fmerge(v, v) &= v \\
\fmerge(n, m) &= {\tt Number} \\
\fmerge(n, v) &= {\tt QData} \\
\fmerge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
\fmerge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
\fmerge({\tt QPair}, v) &= {\tt QData} \\
\vdots
\end{align*}
\caption{Quoted list primitive}
\label{fig:qlist}
\end{figure}

This abstraction was so critical to timely convergence that all of our evaluation uses it.

@@

%% Vincent St Amour for feedback on early drafts.
%% Greg Morrisset and Matthias Felleisen for discussions.

>>>>

\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance and precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } S' &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              F' &= \{c \mid (c,\_) \in S' \setminus S\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) transition elimination
and (2) pragmatic improvement.
%
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item store deltas;
 \item timestamped stores; and
 \item preallocated data structures.
\end{enumerate}

All but the first technique are precision preserving. We will discuss the design rationale
for the move.

@@

\begin{align*}
\force &: \Store \times \Value \to \Set(\Value) \\
\force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
\force(\msto,\mval) &= \{\mval\} \\
\ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto \force(\msto, \mval)] \\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\lmachstep\;
\ev[^\mcntr]{\mexp,\menv,\ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
\mathit{bind\,}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \ext(\msto,\maddr, \mval), \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

@@

\paragraph{Option 1: Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step, since store widening
leads to stepping several states in one big ``step.'' With laziness,
the reference will mean the original binding(s) of the variable or the
new binding, because the actual store lookup is delayed one step
(i.e. laziness is administrative). Without laziness, the reference
will fan out to all the bindings of the variable before the new
binding happens and thus theoretically has an observable precision
difference. % All our benchmarks maintain their precision with lazy non-determinism, however.

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by duplicating the reduction relation to specialize
variable lookup. This works since in the semantics of ISWIM with
store-allocated results consumes the value component of states in one
step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual and implementation complexity for negligible benefit, we
decided against this approach.

@@

\begin{theorem}[Soundness]
  If $\mstate \machstep \mstate'$ and $\mstate \sqsubseteq
  \mastate$ then there exists a $\mastate'$ such that $\mastate
  \lmachstep \mastate'$ and $\mstate' \sqsubseteq \mastate'$
\end{theorem}
Here $\sqsubseteq$ is straightforward - the LHS store must be
contained in the RHS store, and if values occur in the states, forcing
the LHS value must be a subset of forcing the corresponding RHS value.
The proof is by cases on $\mstate \machstep \mstate'$.

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \compile{\mexp}(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{align*}
%% CONTINUE
\co{\kmt,\mval,\msto} &\cmachstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr},\mval,\msto} & \cmachstep
\mcomp^\mcntr(\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr})
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \cmachstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \cmachstep
\mcompi0^\mcntr(\menv,\msto,\mkont)
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \cmachstep
\mcompi1^\mcntr(\menv,\msto,\mkont)
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \cmachstep
\mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
\text{ where }\menv',\msto',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
\\
\ap{\mop,\mval,\msto,\mkont} & \cmachstep
\co{\mkont,\mval',\msto} \\
\text{ where }\mkont &\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\end{gather*}
\caption{Abstract abstract machine for compiled ISWIM}
\label{fig:caam}
\end{figure}

\paragraph{Correctness}
The correctness of abstract compilation seems obvious, but it has
never before been rigorously proved. What constitutes correctness in
the case of dropped states, anyway? Applying an abstract bytecode's
function does many ``steps'' in one go, at the end of which, the two
semantics line up again (modulo representation of expressions). This
constitutes the use of a notion of stuttering. We therefore prove a
stuttering bisimulation~\cite{clarke-stuttering} between this
semantics and the previous section's to show precision preservation.

The number of transitions that can occur in succession from an
abstract bytecode is roughly bounded by the amount of expression
nesting in the program. The partial order that defines subexpressions
smaller than the expression in which they occur is well-founded, so we
use this observation to drive our use of a well-founded equivalence
bisimulation (WEB). WEBs are equivalent to the notion of a stuttering
bisimulation, but are more amenable to mechanisation (and thus
rigorous proof). They also only require reasoning over one step of the
reduction relation, which makes the proof concise.

We define a refinement from non-compiled to compiled states by
``committing'' all the actions of an $\ev{}$ state (defined similarly to
$\compile{\_}$, but immediately applies the functions), and
subsequently changing all expressions with their compiled
variants. Since WEBs are for single transition systems, a WEB
refinement is over the disjoint union of our two semantics, and the
equivalence relation we use is just that a state is related to its
refined state (and itself). Call this relation $B$.

Before we prove this setup is indeed a WEB, we need one lemma that
applying an abstract bytecode's function is equal to refining the
corresponding $\ev{}$ state:
%
\begin{lemma}[Compile/commit]
For all $\mcntr,\mexp,\menv,\msto,\mkont$,
$\compile{\mexp}^\mcntr(\menv,r(\msto),r(\mkont)) = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
\end{lemma}
The proof is by induction on $\mexp$.

\begin{theorem}
$B$ is a WEB on $\lmachstep \cup \cmachstep$
\end{theorem}

The full proof is in the appendix, but it follows by cases on
$\lmachstep \cup \cmachstep$ with the WEB \emph{witness} being the
well-order on expressions (with a $\bot$ element), and the following
$\erankt$, $\erankl$ functions:

\begin{align*}
\erankt(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mexp \\
\erankt(\mstate) &= \bot \quad \text{otherwise} \\
\erankl(s,s') &= 0
\end{align*}
All cases are either simple steps or appeals to the well-order on $\erankt$'s range. The other rank function,
$\erankl$ is unnecessary, so we just make it the constant 0
function. The $\cmachstep$ cases are trivial.

@@

Every step the analysis makes for the above techniques requires
joining entire (large) stores together. Not every step will modify all
addresses of the store, so joining entire stores is wasteful in terms
of memory and time. We can instead log store changes and replay the
change log on the full store after all steps have completed. This uses
far fewer join operations, leading to less overhead, and is \emph{precision-preserving}.

We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
 becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:
% TODO: Turn the recursion here into ...:
\begin{align*}
\replay &: \StoreDelta \times \Store \to \Store \\
\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &= \msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots
\end{align*}
%\replay(\mtlst, \msto) &= \msto \\
%\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])

@@

\begin{align*}
(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \dmachstep
(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
\end{align*}
\begin{align*}
\mathit{bind\,}^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\msdiff}, \mcntr' \\
\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
               \maddr &= \mvar\mcntr'
\end{align*}
Compilation changes to additionally take a $\msdiff$ component, so the
above rule's right hand side would instead be $k^{\mcntr'}(\menv',
\msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.

We also lift $\dmachstep$ to accommodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated:

\begin{align*}
(S, F, \msto) &\mathrel{\damachstep} (S \cup S', F', \msto') \\
\mbox{ where } (F', \msto') &= \step^*(\varnothing, F, \msto) \\
               S' &= \setof{(c,\msto') \mid c \in F'} \\
\step^*(F', \varnothing, \msto) &= (S, \msdiff) \\
\step^*(F', \{c\}\cup F, \msto) &= \step^*(F'\cup \states^*, F, \msto^*) \\
\states^* &= \{c' \mid (c,\msto,\msdiff) \dmachstep (c',\msdiff^c) \} \\
\msto^* &= \replayall(\{\msdiff^c \mid (c, \msto,\msdiff) \dmachstep (c', \msdiff^c)\}, \msto)
\end{align*}

Here $\replayall : \Set(\StoreDelta) \times \Store \to \Store$ replays all logs in a set on top of the given store. Since join is associative and commutative, the order does not matter:
\begin{align*}
  \replayall(\varnothing,\msto) &= \msto \\
  \replayall(\setof{\msdiff} \cup \Xi, \msto) &= \replayall(\Xi, \replay(\msdiff, \msto))
\end{align*}

Correctness here requires a lemma that relates the differences in the compilation functions.
\begin{lemma}
For all $\mexp, \mcntr,\menv,\msto,\msdiff,\mkont,\msto^*,\mkont^*$,
if $\msto \equiv \msto^*$ and $\mkont \equiv \mkont^*$, then
$\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$ iff
$\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \mstate$ such that
$\mastate,\replay(\msdiff',\msto) \equiv \nw(\mstate)$
\end{lemma}
Proof is by induction on $\mexp$.

\begin{theorem}
If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
$(S,F,\msto) \machstep (S',F',\msto')$ iff
$\exists S^*_1, F^*_1, \msto^*_1.
  S' \equiv S^*_1 \wedge F' \equiv F^*_1 \wedge \msto' \equiv \msto^*_1 \wedge
  (S^*,F^*,\msto^*) \dmachstep (S^*_1,F^*_1,\msto^*_1)$
\end{theorem}
Here $\equiv$ means structurally the same, only compiled expressions
are compiled with the updated compilation function that builds store
deltas. The proof follows by equational reasoning and the previous lemma.

@@

@@

%% Vincent St Amour and Mitchell Wand for feedback on early drafts.
%% Greg Morrisset and Matthias Felleisen for discussions.
